---
title: "Differential Expression Analysis: Skeletal Muscle Ageing"
author: "Vishishtaa Pandit"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Introduction

This analysis investigates transcriptomic changes in human skeletal 
muscle across age groups using publicly available RNA-seq data 
(GSE164471, Tumasian et al., Nature Communications 2021).

We compare gene expression between young (20-34 years) and old 
(65+ years) individuals to identify ageing-associated pathways.

## 2. Setup

```{r libraries}
library(DESeq2)
library(tidyverse)
library(pheatmap)
library(clusterProfiler)
library(org.Hs.eg.db)
library(EnhancedVolcano)
```

## 3. Load and Prepare Data

```{r load_data}
# Load count matrix
counts_raw <- read.csv(
  "data/GSE164471_GESTALT_Muscle_ENSG_counts_annotated.csv",
  row.names = 1
)

# Preview
dim(counts_raw)
head(counts_raw[, 1:5])
```

```{r prepare_data}
# Extract only COUNT columns (exclude annotation columns)
count_cols <- grep("_COUNT$", colnames(counts_raw), value = TRUE)
counts <- counts_raw[, count_cols]

# Extract age group from column names
# Column format: MUSCLE_AGE22_M_GROUP20_34_..._COUNT
get_group <- function(colname) {
  if (grepl("GROUP20_34", colname)) return("young")
  if (grepl("GROUP65_79|GROUP80_PLUS", colname)) return("old")
  return(NA)
}

groups <- sapply(colnames(counts), get_group)

# Keep only young and old samples (drop middle-aged)
keep_samples <- !is.na(groups)
counts <- counts[, keep_samples]
groups <- groups[keep_samples]

# Build metadata table
metadata <- data.frame(
  condition = factor(groups, levels = c("young", "old")),
  row.names = colnames(counts)
)

cat("Young samples:", sum(groups == "young"), "\n")
cat("Old samples:", sum(groups == "old"), "\n")
```

## 4. Quality Control

```{r deseq_object}
# Build DESeq2 object
dds <- DESeqDataSetFromMatrix(
  countData = round(counts),  # DESeq2 needs integers
  colData = metadata,
  design = ~ condition
)

# Filter lowly expressed genes (keep genes with >10 counts in >3 samples)
keep <- rowSums(counts(dds) >= 10) >= 3
dds <- dds[keep, ]
cat("Genes after filtering:", nrow(dds), "\n")
```

```{r pca_plot}
# Variance stabilising transformation for QC
vsd <- vst(dds, blind = TRUE)

# PCA plot
pca_plot <- plotPCA(vsd, intgroup = "condition") +
  theme_minimal(base_size = 14) +
  ggtitle("PCA: Young vs Old Skeletal Muscle") +
  scale_color_manual(
    values = c("young" = "#2196F3", "old" = "#F44336")
  )

print(pca_plot)

ggsave("figures/01_pca_plot.png", 
       plot = pca_plot,
       width = 8, height = 6, 
       dpi = 300)
```

## 5. Differential Expression Analysis

```{r run_deseq}
# Run DESeq2
dds <- DESeq(dds)

# Extract results: old vs young
res <- results(dds,
  contrast = c("condition", "old", "young"),
  alpha = 0.05
)

summary(res)
```

```{r results_table}
# Clean results table
res_df <- as.data.frame(res) %>%
  rownames_to_column("gene") %>%
  filter(!is.na(padj)) %>%
  arrange(padj)

# Significant DEGs
sig_genes <- res_df %>%
  filter(padj < 0.05, abs(log2FoldChange) > 1)

cat("Upregulated in old:", sum(sig_genes$log2FoldChange > 0), "\n")
cat("Downregulated in old:", sum(sig_genes$log2FoldChange < 0), "\n")

head(sig_genes, 20)
```

## 6. Visualisation

```{r volcano_plot, fig.width=10, fig.height=8}
volcano <- EnhancedVolcano(res_df,
  lab = res_df$gene,
  x = "log2FoldChange",
  y = "padj",
  title = "Old vs Young Skeletal Muscle",
  subtitle = "DESeq2 | padj < 0.05 | |LFC| > 1",
  pCutoff = 0.05,
  FCcutoff = 1,
  col = c("grey70", "grey70", "#2196F3", "#F44336"),
  colAlpha = 0.6,
  legendPosition = "bottom",
  labSize = 3
)

print(volcano)

ggsave("figures/02_volcano_plot.png",
       plot = volcano,
       width = 10, height = 8,
       dpi = 300)
```

```{r heatmap, fig.width=10, fig.height=12}
top40 <- sig_genes %>%
  arrange(padj) %>%
  head(40) %>%
  pull(gene)

mat <- assay(vsd)[top40, ]
mat_scaled <- t(scale(t(mat)))

annotation_col <- data.frame(
  Condition = metadata$condition,
  row.names = rownames(metadata)
)

ann_colors <- list(
  Condition = c(young = "#2196F3", old = "#F44336")
)

# Save heatmap to file
png("figures/03_heatmap.png", 
    width = 10, height = 12, 
    units = "in", res = 300)

pheatmap(mat_scaled,
  annotation_col = annotation_col,
  annotation_colors = ann_colors,
  show_colnames = FALSE,
  cluster_cols = TRUE,
  color = colorRampPalette(c("#2196F3", "white", "#F44336"))(100),
  main = "Top 40 DEGs: Old vs Young Skeletal Muscle",
  fontsize_row = 8
)

dev.off()

# Also display in report
pheatmap(mat_scaled,
  annotation_col = annotation_col,
  annotation_colors = ann_colors,
  show_colnames = FALSE,
  cluster_cols = TRUE,
  color = colorRampPalette(c("#2196F3", "white", "#F44336"))(100),
  main = "Top 40 DEGs: Old vs Young Skeletal Muscle",
  fontsize_row = 8
)
```

## 7. Pathway Enrichment

```{r go_enrichment, fig.width=10, fig.height=8}
# Upregulated genes in old muscle
up_genes <- sig_genes %>%
  filter(log2FoldChange > 0) %>%
  pull(gene)

# Downregulated genes in old muscle  
down_genes <- sig_genes %>%
  filter(log2FoldChange < 0) %>%
  pull(gene)

# GO enrichment - upregulated
ego_up <- enrichGO(
  gene = up_genes,
  OrgDb = org.Hs.eg.db,
  keyType = "ENSEMBL",
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  readable = TRUE
)

if (nrow(as.data.frame(ego_up)) > 0) {
  p_up <- dotplot(ego_up, showCategory = 15,
    title = "GO BP — Upregulated in Aged Muscle")
  print(p_up)
  ggsave("figures/04_GO_upregulated.png",
         plot = p_up,
         width = 10, height = 8,
         dpi = 300)
}
```

```{r go_down, fig.width=10, fig.height=8}
# GO enrichment - downregulated
ego_down <- enrichGO(
  gene = down_genes,
  OrgDb = org.Hs.eg.db,
  keyType = "ENSEMBL",
  ont = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  readable = TRUE
)

if (nrow(as.data.frame(ego_down)) > 0) {
  p_down <- dotplot(ego_down, showCategory = 15,
    title = "GO BP — Downregulated in Aged Muscle")
  print(p_down)
  ggsave("figures/05_GO_downregulated.png",
         plot = p_down,
         width = 10, height = 8,
         dpi = 300)
}
```


## 8. GenAge Cross-Reference

```{r genage}
# Download GenAge from: https://genomics.senescence.info/genes/human.zip
# Save the CSV as data/genage_human.csv first
# Then uncomment and run:

# genage <- read.csv("data/genage_human.csv")
# genage_genes <- genage$HGNC.symbol
# 
# overlap <- sig_genes %>%
#   filter(gene %in% genage_genes)
# 
# cat("DEGs overlapping with GenAge:", nrow(overlap), "\n")
# 
# ggplot(overlap, 
#        aes(x = log2FoldChange, y = -log10(padj), label = gene)) +
#   geom_point(aes(color = log2FoldChange > 0), size = 3) +
#   geom_text(nudge_y = 0.05, size = 3) +
#   scale_color_manual(
#     values = c("TRUE" = "#F44336", "FALSE" = "#2196F3"),
#     labels = c("Downregulated", "Upregulated")
#   ) +
#   theme_minimal() +
#   labs(title = "GenAge genes differentially expressed with ageing")
```

## 9. Summary

```{r summary}
cat("=== Analysis Summary ===\n")
cat("Dataset: GSE164471 - Skeletal Muscle Transcriptome in Healthy Aging\n")
cat("Young samples (20-34 yrs):", sum(metadata$condition == "young"), "\n")
cat("Old samples (65+ yrs):", sum(metadata$condition == "old"), "\n")
cat("Genes tested:", nrow(res_df), "\n")
cat("Significant DEGs (padj<0.05, |LFC|>1):", nrow(sig_genes), "\n")
cat("Upregulated in aged:", sum(sig_genes$log2FoldChange > 0), "\n")
cat("Downregulated in aged:", sum(sig_genes$log2FoldChange < 0), "\n")
```

## 10. Saving the DEG results in CSV

```{r save_results}
# Save significant DEGs table
write.csv(sig_genes, 
          "results/significant_DEGs.csv", 
          row.names = FALSE)

# Save full results table (all tested genes, not just significant ones)
write.csv(res_df,
          "results/full_DESeq2_results.csv",
          row.names = FALSE)
```